Distortion-corrected kernel density estimate for Riemannian manifolds

### 2022-07-06 paper/paper.Rmd
Change title

---
title: "Kernel density estimate for embedding of Riemannian manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings. @Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point. To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric. The outliers could then be defined as the points with the lowest density estimates. The proposed schematic is shown in Figure \@ref(fig:vkde). The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in
variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.

@@

For the exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives 
$$
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
$$
in normal coordinates. The volume density function is motivated by how a measure $du$ changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
@Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of q, $\exp_q^{-1}(p)$.

@@

## Distotion-corrected kernel density estimator (DC-KDE) {#dckde}

However, in application, we usually find the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$. Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$. Therefore, based on the estimator in \@ref(eq:denriem), we propose a kernel density estimator of $f$ with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be 
\begin{equation}
\label{eq:denestimator}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation} 
where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension $d\times d$.

@@

>>>>

---
title: "Distortion-corrected kernel density estimate for Riemannian manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings. @Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point. To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric. The outliers could then be defined as the points with the lowest density estimates.

@@

For the exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives 
$$
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
$$
in normal coordinates. The volume density function is motivated by how a measure $du$ changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
@Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of q, $\exp_q^{-1}(p)$. 
Note that this is true only when using the geodesic normal coordinates. If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise. Therefore, the Riemannian manifold density estimator in @Pelletier2005-vu only works when the true manifold $M$ is known.

@@

## Distortion-corrected kernel density estimator for manifold embedding {#dckde}

However, in application, we usually find the observations lying on a $d$-dimensional manifold embedded in a much higher $s$-dimensional space, where $d \ll s$. Instead of sampling from the undering manifold $M$, we apply manifold learning algorithms to reduce dimension and find a $d$-dimensional representation of the manifold $M$.
Notice that this representation is not necessarily isometry to $M$, and the underlying manifold is usually unknown in real data sets. Then the density estimator in Equation \@ref(eq:denriem) proposed by @Pelletier2005-vu no longer applies because of the unknown geodesic distance $d_g$ and the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$.
Motivated on the estimator in \@ref(eq:denriem), we propose a distortion-corrected kernel density estimator (DC-KDE) of $f$ with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be 
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r} \frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|} K\bigg( \frac{(\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})}{r} \bigg),
\end{equation} 
where $h(\pmb{y_i}), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and $h(x)$ is positive definite with dimension $d\times d$, $\frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})$ is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point $\pmb{y_i} \ \mathbb{R}^d$. Then the only problem left is the estimation of the Riemannian metric at each embedding point.

@@

The proposed schematic is shown in Figure \@ref(fig:vkde). The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in
variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.
